Linting Linearity in Core/System FC

### 2023-01-19 report.tex
typeset rules and code

Under the lens of a linear type system, consider the variable $p$ to be a linear
resource created by the \textbf{malloc} keyword. Because $p$ is a linear
variable, it must be used exactly once. However, we are using it in two
different calls to \textbf{free}. With our new linearity assumptions, the
% TODO:
example program \emph{would not typecheck}! Some conclusion on linearity
goodness. In section~\ref{linear-types} we describe the fundamentals of linear
types in more detail with a formal definition and many examples.

@@

To understand how linear types are defined and used in practice, we present two
examples of anonymous functions that receive as an argument a handle to work
with and are responsible for closing it, we try to understand how the examples
could be disregarded as incorrect, and work our way up to linear types from
them. The first function ignores the received file handle and returns $\star$
(read unit), which is equivalent to C's \textbf{void}.
% TODO: Better code formatting
\begin{mathpar}
    \lambda h.~\textrm{return}~\star;
    \and
    \lambda h.~\textrm{close}~h;~\textrm{close}~h;
\end{mathpar}
% TODO: Equation references?

Ignoring the file handle which should have been closed by the function makes the
first function incorrect. Similarly, the second function receives the file handle and
closes it twice, which is incorrect not because it falls short of the
specification, but rather because the program will crash while executing it.
%
Additionally, both functions share the same type, $\textbf{Handle} \to \star$, i.e.
a function that takes a \textbf{Handle} and returns $\star$. The second function also
shares this type because \textbf{close} has type $\textbf{Handle} \to \star$.
%
Under a simple type system such as C's both functions are type correct (the
compiler will happily succeed), but both have erroneous behaviours. The first
forgets to close the handle and the second closes it twice. Our goal is now to
reach a type system that rejects these two programs.

The key observation to invalidating these programs is to focus on the function
type going between \textbf{Handle} and $\star$ and augment it to indicate that
\emph{the argument must be used exactly once}, or, in other words, that the
argument of the function must be linear. We take the function type $A \to B$
and replace the function arrow ($\to$) with the linear function arrow ($\lolli$)
%
\footnote{Since linear types are born from a correspondence with linear
logic\cite{linear-logic} (the Curry-Howard isomorphism\cite{curry-howard}), we
borrow the $\lolli$ symbol and other linear logic connectives to describe linear
types.}
%
operator to denote a function that uses its argument exactly once: $A \lolli B$.
%
Providing the more restrictive linear function signature $\textbf{Handle} \lolli
\star$ to the example programs would make both of them fail to typecheck because
they do not satisfy the linearity specification that the function argument
should only be used exactly once.

@@

\begin{mathpar}
  \begin{array}{lcl}
    s,t & ::= & \star\\
        & \mid & t \lolli s\\
        & \mid & t \oplus s\\
        & \mid & t \tensor s\\
        & \mid & t \with s\\
        & \mid & \bang t\\
   \end{array}
\and
  \begin{array}{lcl}
      M,N & ::= & \star \mid \llet{\star = M}{N} \\
        & \mid & u \\
        & \mid & \lambda u. M \mid M~N\\
        & \mid & \textrm{inl}~M \mid \textrm{inr}~M \mid
        \ccase{M}{\textrm{inl}~u_1 \to N_1 \mid \textrm{inr}~u_2 \to N_2}\\
        & \mid & M \tensor N \mid \llet{u_1 \tensor u_2 = M}{N}\\
        & \mid & M \with N \mid \textrm{fst}~M \mid \textrm{snd}~M \\
        & \mid & \bang M \mid \llet{!u = M}{N} \\
   \end{array}
\end{mathpar}

\[
  \begin{array}{c}
    \infer*[right=($\tensor I$)]
    {\Delta \vdash M : A \and \Delta' \vdash N : B}
    {\Delta , \Delta' \vdash (M \tensor N) : A \tensor B}
\qquad
    \infer*[right=($\tensor E$)]
    {\Delta \vdash M : A \tensor B \and \Delta', u{:}A, v{:}B \vdash N : C}
    {\Delta , \Delta' \vdash \llet{u \tensor v}{N} : C}
\\[1em]
    \infer*[right=($\lolli I$)]
    {\Delta , u{:}A \vdash M : B}
    {\Delta \vdash \lambda u. M : A \lolli B}
\qquad
    \infer*[right=($\lolli E$)]
    {\Delta \vdash M : A \lolli B \and \Delta' \vdash N : A}
    {\Delta, \Delta' \vdash M~N : B}
\\[1em]
    \infer*[right=($\with I$)]
    {\Delta \vdash M : A \and \Delta \vdash N : B}
    {\Delta \vdash M \with N : A \with B}
    \qquad
    \infer*[right=($\with E_L$)]
    {\Delta \vdash M : A \with B}
    {\Delta \vdash \textrm{fst}~M : A}
\qquad
    \infer*[right=($\with E_R$)]
    {\Delta \vdash M : A \with B}
    {\Delta \vdash \textrm{snd}~M : B}
\\[1em]
    \infer*[right=($\oplus I_L$)]
    {\Delta \vdash M : A}
    {\Delta \vdash \textrm{inl}~M : A \oplus B}

\qquad
    \infer*[right=($\oplus I_R$)]
    {\Delta \vdash M : B}
    {\Delta \vdash \textrm{inr}~M : A \oplus B}
\\[1em]

    \infer*[right=($\oplus E$)]
    {\Delta \vdash M : A \oplus B \and \Delta', w_1{:}A \vdash N_1 : C \and
    \Delta', w_2{:}B \vdash N_2 : C}
    {\Delta, \Delta' \vdash \ccase{M}{\textrm{inl}~w_1 \to N_1 \mid \textrm{inr}~w_2 \to N_2} : C}
    \\[1em]
    \infer*[right=($\star I$)]
    { }
    {\Delta \vdash \star : \star}
\quad    \infer*[right=($\star E$)]
    {\Delta \vdash M : \star \and \Delta' \vdash N : B}
    {\Delta, \Delta' \vdash \llet{\star = M}{N} : B}
\\[1em]
    \infer*[right=($\bang I$)]
    {\Gamma ; \cdot \vdash M : A}
    {\Gamma ; \cdot \vdash !M : !A}
\qquad    \infer*[right=($\bang E$)]
    {\Gamma ; \Delta \vdash M : \bang A \and \Gamma, u{:}A ; \Delta' \vdash N : C}
    {\Gamma ; \Delta, \Delta' \vdash \llet{!u = M}{N} : C}
  \end{array}
  \]

@@

Haskell is a large language but its small core is based on a typed lambda
calculus. As such, there exist no statements and computation is done simply
through the evaluation of functions. Besides functions, one can define types and
their constructors and pattern match on said constructors. Function application
is denoted by an empty space (\textbf{f~a} means \textbf{f} applied to
\textbf{a}) and pattern matching is done with the \textbf{case} keyword followed
by the enumerated alternatives. All variable names start with lower case and
types start with upper case (excluding type variables). To make explicit the
type of an expression, the \textbf{::} operator is used (e.g.
$\textbf{f~::~Int}\to \textbf{Bool}$ is read \textbf{f} \emph{has type} function
from \textbf{Int} to \textbf{Bool}).

Because Haskell is a pure programming language, input/ouptut side-effects are
"made pure" by modelling them on the type level through the
non-nullary\footnote{\textbf{IO} has kind $\textbf{Type}\to\textbf{Type}$,
that is, it is only a type after another type is passed as a parameter (e.g.
\textbf{IO~Int}, \textbf{IO~Bool}); \textbf{IO} by itself is a \emph{type
constructor}} type constructor \textbf{IO}. Some of the example programs will look though as
if they had statements, but, in reality, the sequential appearence is just
syntatic sugar to an expression using monadic operators. The main take away is
that computations that do I/O may be sequenced together with other operations
that do I/O while retaining the lack of statements and the language purity
guarantees.

@@

Another defining feature of Haskell is its powerful type system. Features such
as type classes, arbitrary-rank types, functional dependencies, GADTs, type
families and (more recently) even impredicative polymorphism\cite{10.1145/3408971} have become
commonplace in Haskell code, enforcing application level invariants and program
correctness through the types. For example, consider the definition of
\textbf{head} in the standard library, a function which takes the first element
of a list by pattern matching on the list constructors.

@@

Note that if the list is empty, we terminate the program with an error. This
function is unsafe -- our program might crash if we use it on an invalid input.
A common technique in Haskell is to use more expressive types to assert
properties about the values and get rid of the unsafe cases (e.g. a
\textbf{NonEmpty} is a list that can not be empty). A well liked motto is "make
invalid states unrepresentable". In this light, we will introduce Generalized
Algebraic Data Types (GADTs) and create a list type indexed by size for which we
can write a completely safe \textbf{head} function. This is done by expressing that the size
of the list must be at least one at the type level.

@@

First, we define the natural numbers inductively: a natural number is either
zero (\textbf{Z}) or a successor (\textbf{S}) of another natural number (e.g.
the successor of the successor of zero is a natural number). The following
definition creates the \textbf{Nat} type, the \textbf{Z} and \textbf{S} data
constructors which construct values, \emph{and} the \textbf{Z} and \textbf{S}
\emph{type} constructors which construct types. The latter two exist because of
the \emph{DataKinds} language feature which promotes term-level constructors to
type-level constructors.

@@

Secondly, we define the data type in GADT syntax for length-index lists which
takes two type parameters. The first type parameter is the length of the list
and the type of the type parameter (i.e. the kind of the first type parameter)
is \textbf{Nat}. To construct a type of kind \textbf{Nat} we can only use the
type constructors \textbf{Z} and \textbf{S}. The second type parameter is the
type of the values contained in the list, and any type is valid, hence the
\textbf{Type} kind.

@@

The length-indexed list is defined inductively as either an empty list \emph{of
size zero}, or the construction of a list by appending a new element to an
existing list \emph{of size $m$} whose final size is $m+1$ (\textbf{S m}). The
list \textbf{Cons~'a'~(Cons~'b'~Nil)} has type \textbf{Vec~(S~(S~Z))~Char}
because \textbf{Nil~::~Vec~Z~Char} and \textbf{Cons~'a'~Nil~::~Vec~(S~Z)~Char}.
GADTs make possible different data constructors being parametrized over
different type parameters as we do with \textbf{Vec}'s size parameter being
different depending on the constructor that constructs the list.

To define the safe \textbf{head} function, we must specify the type of the input
list taking into account that the size must not be zero. To that effect, the
function takes as input a \textbf{Vec (S n) a}, that is, a vector with size
(n+1) for all possible n's, which makes it impossible to pass a list of type
\textbf{Vec Z a} to \textbf{head}.

@@

Pattern matching on the \textbf{Nil} constructor would result in an innacessible
pattern warning because it is impossible to pass a list that would match
\textbf{Nil} -- it would entail the argument list has size zero; however, it's
impossible to call \textbf{head} on a list of size zero!

@@

% A function to compute the length of a list in Haskell is defined by pattern
% matching on the list and returning $0$ if the list is empty and adding one to
% the (computed recursively) length of the rest of the list otherwise. The
% following snippet defines the said \textbf{length} function.
% \begin{code}
%     length [] = 0
%     length (x:xs) = 1 + length xs
% \end{code}
% The function \textbf{length} receives a list and returns an integer, and since
% it treats all list's elements the same, it is said to be polymorphic over the
% list's elements type. In Haskell, function signatures typically appear above the
% definition, and for this particular function we'd have
% \begin{code}
%     length :: [a] -> Int
% \end{code}
% where \textbf{a} is a universally quantified type variable, that is,
% \textbf{length} can be called on lists of ints, lists of characters, lists of
% lists of strings, lists of anything.

% The laziness evaluation strategy entails expressions are only evaluated when
% needed. If we were to call just the above function on a list of computationally
% expensive elements, such as
% \begin{code}
%     length [5^1234, 6^4567, 8^(9^42)]
% \end{code}
% We would get the result 3 instantaneously: the elements of the list would never
% be evaluated because the \textbf{length} function does not require the value of
% each element to compute a result.

>>>>

Under the lens of a linear type system, consider the variable $p$ to be a linear
resource created by the \texttt{malloc} keyword. Because $p$ is a linear
variable, it must be used exactly once. However, we are using it in two
different calls to \texttt{free}. With our new linearity assumptions, the
% TODO:
example program \emph{would not typecheck}! Some conclusion on linearity
goodness. In section~\ref{linear-types} we describe the fundamentals of linear
types in more detail with a formal definition and many examples.

@@

To understand how linear types are defined and used in practice, we present two
examples of anonymous functions that receive as an argument a handle to work
with and are responsible for closing it, we try to understand how the examples
could be disregarded as incorrect, and work our way up to linear types from
them. The first function ignores the received file handle and returns $\star$
(read unit), which is equivalent to C's \texttt{void}.
% TODO: Better code formatting
\begin{mathpar}
    \lambda h.~\textrm{return}~\star;
    \and
    \lambda h.~\textrm{close}~h;~\textrm{close}~h;
\end{mathpar}
% TODO: Equation references?

Ignoring the file handle which should have been closed by the function makes the
first function incorrect. Similarly, the second function receives the file handle and
closes it twice, which is incorrect not because it falls short of the
specification, but rather because the program will crash while executing it.
%
Additionally, both functions share the same type, $\texttt{Handle} \to \star$, i.e.
a function that takes a \texttt{Handle} and returns $\star$. The second function also
shares this type because \texttt{close} has type $\texttt{Handle} \to \star$.
%
Under a simple type system such as C's both functions are type correct (the
compiler will happily succeed), but both have erroneous behaviours. The first
forgets to close the handle and the second closes it twice. Our goal is now to
reach a type system that rejects these two programs.

The key observation to invalidating these programs is to focus on the function
type going between \texttt{Handle} and $\star$ and augment it to indicate that
\emph{the argument must be used exactly once}, or, in other words, that the
argument of the function must be linear. We take the function type $A \to B$
and replace the function arrow ($\to$) with the linear function arrow ($\lolli$)
%
\footnote{Since linear types are born from a correspondence with linear
logic\cite{linear-logic} (the Curry-Howard isomorphism\cite{curry-howard}), we
borrow the $\lolli$ symbol and other linear logic connectives to describe linear
types.}
%
operator to denote a function that uses its argument exactly once: $A \lolli B$.
%
Providing the more restrictive linear function signature $\texttt{Handle} \lolli
\star$ to the example programs would make both of them fail to typecheck because
they do not satisfy the linearity specification that the function argument
should only be used exactly once.

@@

\begin{figure}[h]
\[
\begin{array}{c}
  \begin{array}{lcl}
    A,B & ::= & \star\\
        & \mid & A \lolli B\\
        & \mid & A \oplus B\\
        & \mid & A \tensor B\\
        & \mid & A \with B\\
        & \mid & \bang A\\
   \end{array}
\qquad
  \begin{array}{lcl}
      M,N & ::= & \star \mid \llet{\star = M}{N} \\
        & \mid & u \\
        & \mid & \lambda u. M \mid M~N\\
        & \mid & \textrm{inl}~M \mid \textrm{inr}~M \mid
        \ccase{M}{\textrm{inl}~u_1 \to N_1;\textrm{inr}~u_2 \to N_2}\\
        & \mid & M \tensor N \mid \llet{u_1 \tensor u_2 = M}{N}\\
        & \mid & M \with N \mid \textrm{fst}~M \mid \textrm{snd}~M \\
        & \mid & \bang M \mid \llet{!u = M}{N} \\
   \end{array}
\end{array}
\]
\caption{Grammar for a linearly-typed lambda calculus}
\end{figure}

\begin{figure}[h]
\[
  \begin{array}{c}
    \infer*[right=($\tensor I$)]
    {\Gamma ; \Delta \vdash M : A \and \Gamma ; \Delta' \vdash N : B}
    {\Gamma ; \Delta , \Delta' \vdash (M \tensor N) : A \tensor B}
\qquad
    \infer*[right=($\tensor E$)]
    {\Gamma ; \Delta \vdash M : A \tensor B \and \Gamma ; \Delta', u{:}A, v{:}B \vdash N : C}
    {\Gamma ; \Delta , \Delta' \vdash \llet{u \tensor v}{N} : C}
\\[1em]
    \infer*[right=($\lolli I$)]
    {\Gamma ; \Delta , u{:}A \vdash M : B}
    {\Gamma ; \Delta \vdash \lambda u. M : A \lolli B}
\qquad
    \infer*[right=($\lolli E$)]
    {\Gamma ; \Delta \vdash M : A \lolli B \and \Gamma ; \Delta' \vdash N : A}
    {\Gamma ; \Delta, \Delta' \vdash M~N : B}
\\[1em]
    \infer*[right=($\with I$)]
    {\Gamma ; \Delta \vdash M : A \and \Gamma ; \Delta \vdash N : B}
    {\Gamma ; \Delta \vdash M \with N : A \with B}
    \qquad
    \infer*[right=($\with E_L$)]
    {\Gamma ; \Delta \vdash M : A \with B}
    {\Gamma ; \Delta \vdash \textrm{fst}~M : A}
\qquad
    \infer*[right=($\with E_R$)]
    {\Gamma ; \Delta \vdash M : A \with B}
    {\Gamma ; \Delta \vdash \textrm{snd}~M : B}
\\[1em]
    \infer*[right=($\oplus I_L$)]
    {\Gamma ; \Delta \vdash M : A}
    {\Gamma ; \Delta \vdash \textrm{inl}~M : A \oplus B}
\qquad
    \infer*[right=($\oplus I_R$)]
    {\Gamma ; \Delta \vdash M : B}
    {\Gamma ; \Delta \vdash \textrm{inr}~M : A \oplus B}
\\[1em]
    \infer*[right=($\oplus E$)]
    {\Gamma ; \Delta \vdash M : A \oplus B \and \Gamma ; \Delta', w_1{:}A \vdash N_1 : C \and
    \Gamma ; \Delta', w_2{:}B \vdash N_2 : C}
    {\Gamma ; \Delta, \Delta' \vdash \ccase{M}{\textrm{inl}~w_1 \to N_1 \mid \textrm{inr}~w_2 \to N_2} : C}
    \\[1em]
    \infer*[right=($\star I$)]
    { }
    {\Gamma ; \Delta \vdash \star : \star}
\quad    \infer*[right=($\star E$)]
    {\Gamma ; \Delta \vdash M : \star \and \Gamma ; \Delta' \vdash N : B}
    {\Gamma ; \Delta, \Delta' \vdash \llet{\star = M}{N} : B}
\\[1em]
    \infer*[right=($\bang I$)]
    {\Gamma ; \cdot \vdash M : A}
    {\Gamma ; \cdot \vdash !M : !A}
\qquad    \infer*[right=($\bang E$)]
    {\Gamma ; \Delta \vdash M : \bang A \and \Gamma, u{:}A ; \Delta' \vdash N : C}
    {\Gamma ; \Delta, \Delta' \vdash \llet{!u = M}{N} : C}
  \end{array}
\]
\caption{Typing rules for a linearly-typed lambda calculus}
\end{figure}

@@

Haskell is a large language but its small core is based on a typed lambda
calculus. As such, there exist no statements and computation is done simply
through the evaluation of functions. Besides functions, one can define types and
their constructors and pattern match on said constructors. Function application
is denoted by an empty space (\texttt{f~a} means \texttt{f} applied to
\texttt{a}) and pattern matching is done with the \texttt{case} keyword followed
by the enumerated alternatives. All variable names start with lower case and
types start with upper case (excluding type variables). To make explicit the
type of an expression, the \texttt{::} operator is used (e.g.
$\texttt{f~::~Int}\to \texttt{Bool}$ is read \texttt{f} \emph{has type} function
from \texttt{Int} to \texttt{Bool}).

Because Haskell is a pure programming language, input/ouptut side-effects are
"made pure" by modelling them on the type level through the
non-nullary\footnote{\texttt{IO} has kind $\texttt{Type}\to\texttt{Type}$,
that is, it is only a type after another type is passed as a parameter (e.g.
\texttt{IO~Int}, \texttt{IO~Bool}); \texttt{IO} by itself is a \emph{type
constructor}} type constructor \texttt{IO}. Some of the example programs will look though as
if they had statements, but, in reality, the sequential appearence is just
syntatic sugar to an expression using monadic operators. The main take away is
that computations that do I/O may be sequenced together with other operations
that do I/O while retaining the lack of statements and the language purity
guarantees.

@@

Another defining feature of Haskell is its powerful type system. Features such
as type classes, arbitrary-rank types, functional dependencies, GADTs, type
families and (more recently) even impredicative polymorphism\cite{10.1145/3408971} have become
commonplace in Haskell code, enforcing application level invariants and program
correctness through the types. For example, consider the definition of
\texttt{head} in the standard library, a function which takes the first element
of a list by pattern matching on the list constructors.

@@

Note that if the list is empty, we terminate the program with an error. This
function is unsafe -- our program might crash if we use it on an invalid input.
A common technique in Haskell is to use more expressive types to assert
properties about the values and get rid of the unsafe cases (e.g. a
\texttt{NonEmpty} is a list that can not be empty). A well liked motto is "make
invalid states unrepresentable". In this light, we will introduce Generalized
Algebraic Data Types (GADTs) and create a list type indexed by size for which we
can write a completely safe \texttt{head} function. This is done by expressing that the size
of the list must be at least one at the type level.

@@

First, we define the natural numbers inductively: a natural number is either
zero (\texttt{Z}) or a successor (\texttt{S}) of another natural number (e.g.
the successor of the successor of zero is a natural number). The following
definition creates the \texttt{Nat} type, the \texttt{Z} and \texttt{S} data
constructors which construct values, \emph{and} the \texttt{Z} and \texttt{S}
\emph{type} constructors which construct types.
%
% The latter two exist because of the \emph{DataKinds} language feature which
% promotes term-level constructors to type-level constructors.

@@

Secondly, we define the data type in GADT syntax for length-index lists which
takes two type parameters. The first type parameter is the length of the list
and the type of the type parameter (i.e. the kind of the first type parameter)
is \texttt{Nat}. To construct a type of kind \texttt{Nat} we can only use the
type constructors \texttt{Z} and \texttt{S}. The second type parameter is the
type of the values contained in the list, and any type is valid, hence the
\texttt{Type} kind.

@@

The length-indexed list is defined inductively as either an empty list \emph{of
size zero}, or the construction of a list by appending a new element to an
existing list \emph{of size $m$} whose final size is $m+1$ (\texttt{S m}). The
list \texttt{Cons~'a'~(Cons~'b'~Nil)} has type \texttt{Vec~(S~(S~Z))~Char}
because \texttt{Nil~::~Vec~Z~Char} and \texttt{Cons~'a'~Nil~::~Vec~(S~Z)~Char}.
GADTs make possible different data constructors being parametrized over
different type parameters as we do with \texttt{Vec}'s size parameter being
different depending on the constructor that constructs the list.

To define the safe \texttt{head} function, we must specify the type of the input
list taking into account that the size must not be zero. To that effect, the
function takes as input a \texttt{Vec (S n) a}, that is, a vector with size
(n+1) for all possible n's, which makes it impossible to pass a list of type
\texttt{Vec Z a} to \texttt{head}.

@@

Pattern matching on the \texttt{Nil} constructor would result in an innacessible
pattern warning because it is impossible to pass a list that would match
\texttt{Nil} -- it would entail the argument list has size zero; however, it's
impossible to call \texttt{head} on a list of size zero!

@@

% A function to compute the length of a list in Haskell is defined by pattern
% matching on the list and returning $0$ if the list is empty and adding one to
% the (computed recursively) length of the rest of the list otherwise. The
% following snippet defines the said \texttt{length} function.
% \begin{code}
%     length [] = 0
%     length (x:xs) = 1 + length xs
% \end{code}
% The function \texttt{length} receives a list and returns an integer, and since
% it treats all list's elements the same, it is said to be polymorphic over the
% list's elements type. In Haskell, function signatures typically appear above the
% definition, and for this particular function we'd have
% \begin{code}
%     length :: [a] -> Int
% \end{code}
% where \texttt{a} is a universally quantified type variable, that is,
% \texttt{length} can be called on lists of ints, lists of characters, lists of
% lists of strings, lists of anything.

% The laziness evaluation strategy entails expressions are only evaluated when
% needed. If we were to call just the above function on a list of computationally
% expensive elements, such as
% \begin{code}
%     length [5^1234, 6^4567, 8^(9^42)]
% \end{code}
% We would get the result 3 instantaneously: the elements of the list would never
% be evaluated because the \texttt{length} function does not require the value of
% each element to compute a result.
